Relational Parametricity for Algebraically Indexed Types

### 2012-06-05 paper/section-general-framework.tex
Push the notation changes through, and add some notes

@@

@@

Having defined the language of algebraically-indexed types that we
will consider, we now turn to the denotational semantics of these
types. In \autoref{sec:index-erasure-semantics} we define an
\emph{index-erasure} interpretation of types that interprets every
well-indexed type as a set, ignoring the indexing by algebraic
terms. In the index erasure semantics, the semantics of a well-indexed
type $\tyPrim{X}{e_1,...,e_n}$ is determined solely by the primitive
type $\tyPrimNm{X}$ and not by the index terms $e_1,...,e_n$. Building
on the index-erasure semantics, we define in
\autoref{sec:relational-semantics} the \emph{relational}
interpretation of types.

@@

For each primitive type $\tyPrimNm{X} \in \PrimTypeSet$, we assume
that $\tyPrimNm{X}$ is assigned a set $\tyPrimSem{\tyPrimNm{X}}$. We
extend this to assign a set to every well-formed type by induction on
the type structure by the following clauses.
\begin{displaymath}
  \begin{array}{@{\hspace{0.2em}}c@{\hspace{0em}}c}
    \begin{array}{l}
      \tySem{\tyUnit} \isDefinedAs \{*\} \\
      \tySem{A \tyProduct B} \isDefinedAs \tySem{A} \times \tySem{B} \\
      \tySem{A \tyArr B} \isDefinedAs \tySem{A} \to \tySem{B} \\
    \end{array}
    &
    \begin{array}{l}
      \tySem{A + B} \isDefinedAs \tySem{A} + \tySem{B} \\
      \tySem{\tyPrim{X}{e_1,...,e_n}} \isDefinedAs \tyPrimSem{\tyPrimNm{X}} \\
      \tySem{\forall i\mathord:s. A} \isDefinedAs \tySem{A}
    \end{array}
  \end{array}
\end{displaymath}
The interpretation of the unit type is a chosen one element set, and
the interpretations of the function, tuple and sum types is simply by
the corresponding construction on sets. The interpretation of the
universal quantifier simply ignores the indexing: the interpretation
of the type $\forall i\mathord:s.A$ is exactly the interpretation of
the type $A$.

\begin{example}[Units of Measure] Kennedy's original units of measure
  system seen as an instantiation of our system uses the assignment
  $\tyPrimSem{\mathsf{unit}} \isDefinedAs \mathbb{Q}$, where
  $\mathbb{Q}$ is the set of rational numbers.
\end{example}

Since the index-erasure interpretation has completely ignores the
index expressions and type equality is defined as an extension of
index equality, it is straightforward to prove that equal types have
equal denotations when interpreted in the index-erasure semantics, and
that substitution of index terms has no effect on the index-erasure
interpretation of types:
\begin{lemma}\label{lem:tyeqsubst-erasure}
  \begin{enumerate}
  \item If $\Delta \vdash A \equiv B \isType$ then $\tySem{A} =
    \tySem{B}$; and
  \item If $\Delta \vdash A \isType$ and $\sigma : \Delta' \Rightarrow
    \Delta$, then $\tySem{\sigma^*A} = \tySem{A}$.
  \end{enumerate}
\end{lemma}

@@

For the unit, tuple, sum and function types the we will define the
relational interpretation as a standard logical relation. The
relational interpretations of primitive types with index arguments and
the universally quantified types requires an interpretation of index
contexts. We assign interpretations to index contexts in terms of
\emph{relational environments}, which we now define.

A relation environment for a context $\Delta$ is a
(dependently typed) function that, for each primitive type
$\tyPrimNm{X}$ and instantiation of its arguments $\sigma$, assigns a
binary relation $\rho\ \tyPrimNm{X}\ \sigma$ on the index-erasure
interpretation of $\tyPrimNm{X}$:
\begin{displaymath}
  \rho : (\tyPrimNm{X} \in \PrimTypeSet) \to (\Delta \Rightarrow \primTyArity(\tyPrimNm{X})) \to \Rel(\tyPrimSem{\tyPrimNm{X}})
\end{displaymath}
Every relation environment must respect index term equality: for any
$\tyPrimNm{X} \in \PrimTypeSet$ and pair of simultaneous substitutions
$\sigma : \Delta \Rightarrow \primTyArity(\tyPrimNm{X})$ and $\sigma'
: \Delta \Rightarrow \primTyArity(\tyPrimNm{X})$ such that $\Delta
\vdash \sigma \equiv \sigma' \Rightarrow \primTyArity(\tyPrimNm{X})$, % FIXME: make sure the notation for ss equality is consistent
then $\rho\ \tyPrimNm{X}\ \sigma = \rho\ \tyPrimNm{X}\ \sigma'$.

@@

We write $\mathrm{RelEnv}(\Delta)$ for the set of all relation
environments for a context $\Delta$. Given a relational environment
$\rho \in \mathrm{RelEnv}(\Delta')$ and a simultaneous substitution
$\sigma : \Delta \Rightarrow \Delta'$, we can derive the composed
relational environment (with an abuse of notation) $\rho \circ \sigma
\in \mathrm{RelEnv}(\Delta)$ as $(\rho \circ \sigma)\ \tyPrimNm{X}\
\sigma' = \rho\ \tyPrimNm{X}\ (\sigma \circ \sigma')$.

@@

As we stated above, index contexts $\Delta$ are interpreted as certain
sets of relation environments, i.e.~ subsets of
$\mathrm{RelEnv}(\Delta)$. Depending on the primitive operations that
we assume for our system, we can have different sets of relational
environments that enforce different invariants. For the general
framework, we assume that, for each index context $\Delta$, we are
given a set of relation environments $\relEnv{E}(\Delta)$. Note that
this assignment of sets of relation environments to index contexts is
not assumed to be compositional: it is not necessarily the case that
the set of relational environments $\relEnv{E}(\Delta_1,\Delta_2)$ is
defined in terms of $\relEnv{E}(\Delta_1)$ and
$\relEnv{E}(\Delta_2)$. However, it must satisfy the following two
properties:
\begin{enumerate}
\item Closure under simultaneous substitution of index terms: if $\rho
  \in \relEnv{E}(\Delta)$ and $\sigma : \Delta' \Rightarrow \Delta$,
  then $\rho \circ \sigma \in \relEnv{E}(\Delta')$.
\item If we have a pair of relation environments $\rho_1 \in
  \relEnv{E}(\Delta', i\mathord:s')$ and $\rho_2 \in
  \relEnv{E}(\Delta)$, along with an simultaneous substitution $\sigma
  : \Delta' \Rightarrow \Delta$, such that the outer edges of the
  following diagram commute, for all primitive types $\tyPrimNm{X} \in
  \mathit{PrimType}$:
  %FIXME: check that these compositions are the right way round
  %FIXME: re-do this as a set of equations
  %FIXME: give a better English description of why this is needed
  \begin{displaymath}
    \xymatrix{
      {\Delta \Rightarrow \primTyArity(\tyPrimNm{X})} \ar[r]^(.45){- \circ \pi_{i\mathord:s'}} \ar[d]_{- \circ \sigma}
      &
      {\Delta,i\mathord:s' \Rightarrow \primTyArity(\tyPrimNm{X})} \ar[d]^{- \circ \sigma_{s'}} \ar@/^/[rdd]^{\rho_1 \tyPrimNm{X}}
      \\
      {\Delta' \Rightarrow \primTyArity(\tyPrimNm{X})} \ar[r]^(.45){- \circ \pi_{i\mathord:s'}} \ar@/_/[rrd]_{\rho_2\ \tyPrimNm{X}}
      &
      {\Delta,i\mathord:s' \Rightarrow \primTyArity(\tyPrimNm{X})} \ar@{.>}[dr]_{\rho\ \tyPrimNm{X}}
      \\
      &
      &
      {\Rel(\tyPrimSem{\tyPrimNm{X}})}
    }
  \end{displaymath}
  Then there exists a relation environment $\rho \in
  \relEnv{E}(\Delta,i\mathord:s')$ (the dotted arrow) such that the
  two triangles in the bottom right of the diagram commute.
  % FIXME: mention that we already know that the square commutes
\end{enumerate}

@@

A relational environment $\rho \in \relEnv{E}(\Delta)$ is extended to
a relation interpretation of any type $\Delta \vdash A \isType$ by
induction on its derivation: % Thus, we define a relation
% $\rsem{A}{\relEnv{E}}{\rho} \in \Rel(\tySem{A})$ for each well-indexed
% type.
\begin{eqnarray*}
  \rsem{\tyUnit}{\relEnv{E}}\rho & \isDefinedAs & \{(*,*)\} \\
  \rsem{\tyPrim{X}{e_1,...,e_n}}{\relEnv{E}}\rho & \isDefinedAs & \rho\ {\tyPrimNm{X}}\ (e_1,...,e_n) \\
  \rsem{A \tyArr B}{\relEnv{E}}\rho & \isDefinedAs & \rsem{A}{\relEnv{E}}\rho \relArrow \rsem{B}{\relEnv{E}}\rho \\
  \rsem{A \tyProduct B}{\relEnv{E}}\rho & \isDefinedAs & \rsem{A}{\relEnv{E}}\rho \relTimes \rsem{B}{\relEnv{E}}\rho \\
  \rsem{A + B}{\relEnv{E}}\rho & \isDefinedAs & \rsem{A}{\relEnv{E}}\rho \relSum \rsem{B}{\relEnv{E}}\rho \\
  \rsem{\forall i\mathord:s.A}{\relEnv{E}}\rho & \isDefinedAs & \bigcap\{ \rsem{A}{\relEnv{E}}\rho' \sepbar \rho' \in \extends{\rho}{i\mathord:s} \}
\end{eqnarray*}
In this definition we have made use of the following three
constructions on binary relations. If $R \in \Rel(X)$ and $S \in
\Rel(Y)$, then $R \relArrow S \in \Rel(X \to Y)$ is defined as $\{
(f_1,f_2) \sepbar \forall (a_1,a_2) \in R.\ (f_1a_1,f_2a_2) \in S
\}$. With the same assumptions on $R$ and $S$, the relation $R
\relTimes S \in \Rel(X \times Y)$ is defined as $\{
((a_1,b_1),(a_2,b_2)) \sepbar (a_1,a_2) \in R \land (b_1,b_2) \in S
\}$. Finally, the relation $R \relSum S \in \Rel(X + Y)$ is defined as
$\{ (\mathrm{inl}\ x, \mathrm{inl}\ x') \sepbar (x,x') \in R \} \cup
\{ (\mathrm{inr}\ y, \mathrm{inr}\ y') \sepbar (y,y') \in S \}$.

@@

We now present the rules for well-typed programs over the collection
of types we generated in
\autoref{sec:algebraically-indexed-types}. Each well-type program is
assigned an index-erasure semantics, building on the index-erasure
semantics of types we defined in
\autoref{sec:index-erasure-semantics}. Our main result
(\thmref{thm:abstraction}) is that the index-erasure semantics is
every well-typed program is related to itself in the relational
interpretation of its type: this is the abstraction theorem for every
calculus in our general framework. In the next section we will use
this theorem to derive several interesting results about the programs
expressible in instances of our general framework.

@@

  \inferrule*
  {\Delta \vdash \Gamma \isCtxt \\ \Delta \vdash A \isType \\ x \not\in \Gamma}
  {\Delta \vdash \Gamma, x : A \isCtxt}
\end{mathpar}
Substitution by a simultaneous substitution extends to typing contexts
by performing substitution pointwise on each individual type.

The typing rules for our system define the well-typed terms with
respect to an index context $\Delta$ and a type context $\Delta \vdash
\Gamma \isCtxt$. The judgement $\Delta; \Gamma \vdash M : A$ is
defined in \autoref{fig:programs}. The equational theory on types is
incorporated into the type system via the rule \TirName{TyEq}, which
allows for a term that is judged to have type $A$ to also have any
other equal type $B$.
\begin{figure*}[t]
  \centering
  \begin{mathpar}
    \inferrule* [right=Var]
    {\Delta \vdash \Gamma \isCtxt \\ x : A \in \Gamma}
    {\Delta; \Gamma \vdash x : A}

>>>>

\fixme{Update this introduction}

@@

We write $\Delta \Rightarrow \Delta'$ for the set of all simultaneous
substitutions $\sigma$ such that $\Delta \vdash \sigma \Rightarrow
\Delta'$. When we discuss the relational interpretation of types
below, it will be useful to think of any sequence of sorts as an index
context. \fixme{flesh this out}

@@

Having defined the language of algebraically-indexed types, we now
turn to their denotational interpretation. In
\autoref{sec:index-erasure-semantics} we define an
\emph{index-erasure} interpretation of types that interprets every
well-indexed type as a set, ignoring the indexing expressions. Building on
the index-erasure semantics, we define in
\autoref{sec:relational-semantics} the \emph{relational}
interpretation of types.

@@

The defining feature of the index erasure interpretation is that the
semantics of a well-indexed type $\tyPrim{X}{e_1,...,e_n}$ is
determined solely by the primitive type $\tyPrimNm{X}$ and not by the
index terms $e_1,...,e_n$. Accordingly, we assume that each primitive
type $\tyPrimNm{X} \in \PrimTypeSet$ is assigned a set
$\tyPrimSem{\tyPrimNm{X}}$. We extend this assignment to every
well-indexed type by induction on the type structure:
\begin{displaymath}
  \begin{array}{@{\hspace{0.2em}}c@{\hspace{0em}}c}
    \begin{array}{l}
      \tySem{\tyUnit} \isDefinedAs \{*\} \\
      \tySem{A \tyProduct B} \isDefinedAs \tySem{A} \times \tySem{B} \\
      \tySem{A \tyArr B} \isDefinedAs \tySem{A} \to \tySem{B} \\
    \end{array}
    &
    \begin{array}{l}
      \tySem{A + B} \isDefinedAs \tySem{A} + \tySem{B} \\
      \tySem{\tyPrim{X}{e_1,...,e_n}} \isDefinedAs \tyPrimSem{\tyPrimNm{X}} \\
      \tySem{\forall i\mathord:s. A} \isDefinedAs \tySem{A}
    \end{array}
  \end{array}
\end{displaymath}
The interpretation of the unit type is a chosen one element set, and
the interpretations of the function, tuple and sum types are simply
the corresponding constructions on sets. The interpretation of the
universal quantifier ignores the indexing: the interpretation of the
type $\forall i\mathord:s.A$ is exactly the interpretation of the type
$A$.

\begin{example}[Origin Invariance]
  \fixme{do this}
\end{example}

\begin{example}[Units of Measure] Kennedy's original units of measure
  system seen as an instantiation of our system uses the assignment
  $\tyPrimSem{\mathsf{unit}} \isDefinedAs \mathbb{Q}$, where
  $\mathbb{Q}$ is the set of rational numbers.

  FIXME: Also do the primitive operations
\end{example}

% The index-erasure interpretation completely ignores index expressions
% and type equality is defined as an extension of index
% equality. Therefore, it is straightforward to prove that equal types
% have equal denotations when interpreted in the index-erasure
% semantics, and that substitution of index terms has no effect on the
% index-erasure interpretation of types:
\begin{lemma}\label{lem:tyeqsubst-erasure}
  \begin{enumerate}
  \item If $\Delta \vdash A \equiv B \isType$ then $\tySem{A} =
    \tySem{B}$; and
  \item If $\Delta' \vdash A \isType$ and $\Delta \vdash \sigma
    \Rightarrow \Delta'$, then $\tySem{\sigma^*A} = \tySem{A}$.
  \end{enumerate}
\end{lemma}

@@

For the unit, tuple, sum and function types we will define the
relational interpretation as a standard logical relation. The
relational interpretations of primitive types with index arguments and
the universally quantified types requires an interpretation of index
contexts. We assign interpretations to index contexts in terms of
\emph{relational environments}, which we now define. \fixme{Backwards ref}

A relation environment for a context $\Delta$ is a
(dependently typed) function that, for each primitive type
$\tyPrimNm{X}$ and instantiation of its arguments $\sigma$, assigns a
binary relation $\rho\ \tyPrimNm{X}\ \sigma$ on the index-erasure
interpretation of $\tyPrimNm{X}$:
\begin{displaymath}
  \rho : (\tyPrimNm{X} \in \PrimTypeSet) \to (\Delta \Rightarrow \primTyArity(\tyPrimNm{X})) \to \Rel(\tyPrimSem{\tyPrimNm{X}})
\end{displaymath}
Every relation environment must respect index term equality: for any
$\tyPrimNm{X} \in \PrimTypeSet$ and pair of simultaneous substitutions
$\Delta \vdash \sigma \Rightarrow \primTyArity(\tyPrimNm{X})$ and
$\Delta \vdash \sigma' \Rightarrow \primTyArity(\tyPrimNm{X})$ such
that $\Delta \vdash \sigma \equiv \sigma' \Rightarrow
\primTyArity(\tyPrimNm{X})$, then $\rho\ \tyPrimNm{X}\ \sigma = \rho\
\tyPrimNm{X}\ \sigma'$.

@@

We write $\mathrm{RelEnv}(\Delta)$ for the set of all relation
environments for a context $\Delta$. Given a relational environment
$\rho \in \mathrm{RelEnv}(\Delta')$ and a simultaneous substitution
$\Delta \vdash \sigma \Rightarrow \Delta'$, we can derive the composed
relational environment (with an abuse of notation) $\rho \circ \sigma
\in \mathrm{RelEnv}(\Delta)$ as $(\rho \circ \sigma)\ \tyPrimNm{X}\
\sigma' = \rho\ \tyPrimNm{X}\ (\sigma' \circ \sigma)$. \fixme{Check this is the right way round}

@@

As we stated above, index contexts $\Delta$ are interpreted as certain
sets of relation environments, i.e.~ subsets of
$\mathrm{RelEnv}(\Delta)$. Depending on the primitive operations that
we assume for our system, we may have different sets of relational
environments that enforce different invariants. For the general
framework, we assume that, for each index context $\Delta$, we are
given a set of relation environments $\relEnv{E}(\Delta)$. Note that
this assignment of sets of relation environments to index contexts is
not assumed to be compositional: it is not necessarily the case that
the set of relational environments $\relEnv{E}(\Delta_1,\Delta_2)$ is
defined in terms of $\relEnv{E}(\Delta_1)$ and
$\relEnv{E}(\Delta_2)$. However, it must satisfy the following two
properties:
\begin{enumerate}
\item Closure under simultaneous substitution of index terms: if $\rho
  \in \relEnv{E}(\Delta)$ and $\Delta' \vdash \sigma \Rightarrow
  \Delta$, then $\rho \circ \sigma \in \relEnv{E}(\Delta')$.
\item If we have a pair of relation environments $\rho_1 \in
  \relEnv{E}(\Delta', i\mathord:s')$ and $\rho_2 \in
  \relEnv{E}(\Delta)$, along with an simultaneous substitution $\sigma
  : \Delta' \Rightarrow \Delta$, such that the outer edges of the
  following diagram commute, for all primitive types $\tyPrimNm{X} \in
  \mathit{PrimType}$:
  %FIXME: check that these compositions are the right way round
  %FIXME: re-do this as a set of equations
  %FIXME: give a better English description of why this is needed
  \begin{displaymath}
    \xymatrix{
      {\Delta \Rightarrow \primTyArity(\tyPrimNm{X})} \ar[r]^(.45){- \circ \pi_{i\mathord:s'}} \ar[d]_{- \circ \sigma}
      &
      {\Delta,i\mathord:s' \Rightarrow \primTyArity(\tyPrimNm{X})} \ar[d]^{- \circ \sigma_{s'}} \ar@/^/[rdd]^{\rho_1 \tyPrimNm{X}}
      \\
      {\Delta' \Rightarrow \primTyArity(\tyPrimNm{X})} \ar[r]^(.45){- \circ \pi_{i\mathord:s'}} \ar@/_/[rrd]_{\rho_2\ \tyPrimNm{X}}
      &
      {\Delta,i\mathord:s' \Rightarrow \primTyArity(\tyPrimNm{X})} \ar@{.>}[dr]_{\rho\ \tyPrimNm{X}}
      \\
      &
      &
      {\Rel(\tyPrimSem{\tyPrimNm{X}})}
    }
  \end{displaymath}
  Then there exists a relation environment $\rho \in
  \relEnv{E}(\Delta,i\mathord:s')$ (the dotted arrow) such that the
  two triangles in the bottom right of the diagram commute.
  % FIXME: mention that we already know that the square commutes
\end{enumerate}

@@

We assign a relational interpretation to all well-indexed types
$\Delta \vdash A \isType$ by induction on their derivations,
parameterised by relational environments $\rho \in
\relEnv{E}(\Delta)$:
\begin{eqnarray*}
  \rsem{\tyUnit}{\relEnv{E}}\rho & \isDefinedAs & \{(*,*)\} \\
  \rsem{\tyPrim{X}{e_1,...,e_n}}{\relEnv{E}}\rho & \isDefinedAs & \rho\ {\tyPrimNm{X}}\ (e_1,...,e_n) \\
  \rsem{A \tyArr B}{\relEnv{E}}\rho & \isDefinedAs & \rsem{A}{\relEnv{E}}\rho \relArrow \rsem{B}{\relEnv{E}}\rho \\
  \rsem{A \tyProduct B}{\relEnv{E}}\rho & \isDefinedAs & \rsem{A}{\relEnv{E}}\rho \relTimes \rsem{B}{\relEnv{E}}\rho \\
  \rsem{A + B}{\relEnv{E}}\rho & \isDefinedAs & \rsem{A}{\relEnv{E}}\rho \relSum \rsem{B}{\relEnv{E}}\rho \\
  \rsem{\forall i\mathord:s.A}{\relEnv{E}}\rho & \isDefinedAs & \bigcap\{ \rsem{A}{\relEnv{E}}\rho' \sepbar \rho' \in \extends{\rho}{i\mathord:s} \}
\end{eqnarray*}
In this definition we have made use of the following three
constructions on binary relations. If $R \in \Rel(X)$ and $S \in
\Rel(Y)$, then $R \relArrow S \in \Rel(X \to Y)$ is defined as $\{
(f_1,f_2) \sepbar \forall (a_1,a_2) \in R.\ (f_1a_1,f_2a_2) \in S
\}$. With the same assumptions on $R$ and $S$, the relation $R
\relTimes S \in \Rel(X \times Y)$ is defined as $\{
((a_1,b_1),(a_2,b_2)) \sepbar (a_1,a_2) \in R \land (b_1,b_2) \in S
\}$. Finally, the relation $R \relSum S \in \Rel(X + Y)$ is defined as
$\{ (\mathrm{inl}\ x, \mathrm{inl}\ x') \sepbar (x,x') \in R \} \cup
\{ (\mathrm{inr}\ y, \mathrm{inr}\ y') \sepbar (y,y') \in S \}$.

@@

We now present the rules for well-typed programs over the collection
of types we generated in
\autoref{sec:algebraically-indexed-types}. Each well-type program is
assigned an index-erasure semantics, building on the index-erasure
semantics of types we defined in
\autoref{sec:index-erasure-semantics}. Our main result
(\thmref{thm:abstraction}) is that the index-erasure semantics of
every well-typed program is related to itself in the relational
interpretation of its type: this is the abstraction theorem for every
calculus in our general framework. In the \fixme{next section} we will
use this theorem to derive several interesting results about the
programs expressible in instances of our general framework.

@@

  \inferrule*
  {\Delta \vdash \Gamma \isCtxt \\ \Delta \vdash A \isType \\ x \not\in \Gamma}
  {\Delta \vdash \Gamma, x : A \isCtxt}
\end{mathpar}
Application of simultaneous substitutions extends to typing contexts
by applying the simultaneous substitution to each type.

The typing rules for our system define the well-typed terms with
respect to an index context $\Delta$ and a type context $\Delta \vdash
\Gamma \isCtxt$. The judgment $\Delta; \Gamma \vdash M : A$ is defined
in \autoref{fig:programs}. The equational theory on types is
incorporated into the type system via the rule \TirName{TyEq}, which
allows for a term that is judged to have type $A$ to also have any
other equal type $B$.
\begin{figure*}[t]
  \centering
  \begin{mathpar}
    \inferrule* [right=Var]
    {\Delta \vdash \Gamma \isCtxt \\ x : A \in \Gamma}
    {\Delta; \Gamma \vdash x : A}
